Fail Faster

### 2025-03-05 paper/paper.tex
some things, plus buiild issues

\tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
\subsubsection{Monadic bind}

@@

% \subsubsection{Function call overhead}
\subsubsection{Boxed integers}

@@

\jwc{
  \begin{itemize}
    \item Careful readers might note that the types of return and bind are not actually the right types.
    \item This means that they aren't compatible with ppx-jane syntax.
    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}. This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'.
    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion
  \end{itemize}
}

@@

\begin{lstlisting}
module CodeCps = struct
  type 'a t = 'z. ('a -> 'z code) -> 'z code

  let run (t : 'a code t) : 'a code = t (fun x -> x)

  let let_insert (cx : 'a code) : 'a code t =
    fun k -> k .< let x = .~cx in .~(k .<x>.) >.
end

@@

  let bind (g : 'a t) (f : 'a -> 'b t) =
    fun size random ->
      f (g size random) size random

@@

>>>>

\tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
\subsubsection{Monadic Bind}

@@

% \subsubsection{Function call overhead}
\subsubsection{Boxed RNG}

\jwc{
  \begin{itemize}
    \item The pRNG at the heart of BQ is the hottest part of the hot path: generating a single int with \texttt{int st lo hi} samples from the underlying rng many many times.
    \item BQ's prng is the SplitMix algorithm, implemented in the splittablerandom package.
    \item The precise details of how SplitMix works are unimportant for this paper, but the core of it does an enormous amount of 64-bit int arithmetic.
    \item Because of implementation details related to the garbage collector, OCaml's \texttt{int} type is 63 bits, so an \texttt{Int64.t} is actually \emph{boxed}: a pointer to a custom block, not scanned by the garbage collector.
    \item Because of this, all \texttt{Int64.t} operations (both arithmetic and bitwise operations) must allocate! A single call to \texttt{int} in Bq calls the samples the RNG many times, and each sample (call to \texttt{next\_int64}) allocates 9 words (2 multiplications, one addition, and 6 logical operations).
  \end{itemize}
}

\jwc{We discuss ways to fix this in section ...}

\jwc{We note that this inefficiency is not due to the ``abstraction overhead'' of the DSL itself, but simply due to language implementation and algorithm choices.
We include this to demonstrate the amount of slack in the system, and so we can demonstrate the additional testing-time savings in the evaluation section.
}

@@

\jwc{
  \begin{itemize}
    \item Careful readers might note that the types of return and bind are not actually the right types.
    \item This means that they aren't compatible with ppx-jane syntax.
    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}. This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'. (Footnote: this is an instance of the \emph{codensity} monad, cf ``Asymptotic Improvement of Computations over Free Monads'')
    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
    \item of course, \texttt{codecps} has the usual monad instance for the CPS monad with a polymorphic result type.
    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion, and we get a real monad instance.
  \end{itemize}
}

@@

\begin{lstlisting}
module CodeCps = struct
  type 'a t = { cps : 'z. ('a -> 'z code) -> 'z code }

  let return x = {cps = fun k -> k x}

  let bind (x : 'a t) (f : 'a -> 'b t) : 'b t =
    {cps = fun k -> x.cps (fun a -> (f a).cps k)}

  let run (t : 'a code t) : 'a code = t.cps (fun x -> x)

  let let_insert (cx : 'a code) : 'a code t =
    {cps = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}
end

@@

  let bind (g : 'a t) (f : 'a -> 'b t) =
    fun size random ->
      CodeCps.bind (g size random) (fun x ->
        (f x) size random
      )

@@

\subsection{Faster Random Number Generators}

\jwc{
  \begin{itemize}
    \item 
  \end{itemize}
}
